Illustrating Model Properties

### 2024-06-26 data.qmd
update risk scenarios & switch to SVG

plt.savefig(paths.figure_dir / "fig-data-strat.png", dpi=400)
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-data-upset-ipsi.png", dpi=400)
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-data-upset-contra.png", dpi=400)
plt.show()
```

>>>>

plt.savefig(paths.figure_dir / "fig-data-strat.svg")
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-data-upset-ipsi.svg")
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-data-upset-contra.svg")
plt.show()
```

### 2024-06-26 manuscript.qmd
update risk scenarios & switch to SVG

![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig-data-strat.png){#fig-data-strat}

@@

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=40%}

@@

![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars. Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot. In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.png){#fig-model-midext-evo width="50%"}

@@

![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.png){#fig-model-state-dist}

@@

![Monitoring quantities during the burn-in phase of the parameter sampling. Left: The autocorrelation time of the sampling chain estimated at different sampling steps. We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps. Right: Fraction of accepted MCMC proposals averaged over all parallel walkers. Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.png){#fig-model-burnin-history}

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.png){#fig-model-prevalences-overall}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange. These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}

@@

![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category. While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data. Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.png){#fig-model-prevalences-midext width="50%"}

@@

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and ipsilateral clinical involvement. Unless otherwise mentioned in the legend, the entire contralateral side was assumed to be clinically node negative.](figures/fig-model-risks.png){#fig-model-risks}

@@

![Posterior distributions over the data prevalence for contralateral LNL II involvement under the scenario of early T-category and midline extension. The green beta distribution shows this prevalence for those patients for which midline extension is known. While the blue PDF is a marginal distribution summed over all possible combinations, which are shown in shades of red. The more permutations a particular combination has, the more opaque it is plotted.](figures/fig-model-marginal-pdf.png){#fig-model-marginal-pdf width="50%"}

>>>>

![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig-data-strat.svg){#fig-data-strat}

@@

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/small-graph.svg){#fig-small-graph width=40%}

@@

![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars. Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot. In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.svg){#fig-model-midext-evo width="50%"}

@@

![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.svg){#fig-model-state-dist}

@@

![Monitoring quantities during the burn-in phase of the parameter sampling. Left: The autocorrelation time of the sampling chain estimated at different sampling steps. We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps. Right: Fraction of accepted MCMC proposals averaged over all parallel walkers. Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.svg){#fig-model-burnin-history}

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange. These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.svg){#fig-model-prevalences-with-ipsi}

@@

![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category. While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data. Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.svg){#fig-model-prevalences-midext width="50%"}

@@

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses. All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative.](figures/fig-model-risks.svg){#fig-model-risks}

@@

![Posterior distributions over the data prevalence for contralateral LNL II involvement under the scenario of early T-category and midline extension. The green beta distribution shows this prevalence for those patients for which midline extension is known. While the blue PDF is a marginal distribution summed over all possible combinations, which are shown in shades of red. The more permutations a particular combination has, the more opaque it is plotted.](figures/fig-model-marginal-pdf.svg){#fig-model-marginal-pdf width="50%"}

### 2024-06-26 model.qmd
update risk scenarios & switch to SVG

plt.savefig(paths.figure_dir / "fig-model-midext-evo.png", dpi=400)
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-burnin-history.png", dpi=400)
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-state-dist.png", dpi=400)
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-prevalences-overall.png", dpi=400)
plt.show()
```

@@

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[0,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[1,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.16, 0.99), (0.53, 0.55)],
)
split_legends(
  axes=axes[1,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.13, 0.55), (0.5, 0.99)],
)

plt.savefig(paths.figure_dir / "fig-model-prevalences-with-ipsi.png", dpi=400)
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-prevalences-midext.png", dpi=400)
plt.show()
```

@@

```{python}
import numpy as np
import matplotlib.pyplot as plt
from matplotlib.legend_handler import HandlerTuple
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, draw, COLORS, split_legends
from scripts import paths, shared

all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}

@@

fig, ax = plt.subplots(nrows=nrows, ncols=ncols)

contents = []
defaults = {
  "density": True,
  "histtype": "stepfilled",
  "alpha": 0.7,
  "bins": 50,
}

# contra LNL II
scenario = Scenario(
  t_stages=["early"],
  midext=False,
  diagnosis={
    "ipsi": {"CT": all_false},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
contents.append(Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
  color=COLORS["green"],
  label="early, lateralized; N0",
))

scenario = Scenario(
  t_stages=["late"],
  midext=False,
  diagnosis={
    "ipsi": {"CT": {"III": True}},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
contents.append(Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
  color=COLORS["blue"],
  label="adv., lateralized; ipsi LNL III",
))

scenario = Scenario(
  t_stages=["early"],
  midext=True,
  diagnosis={
    "ipsi": {"CT": all_false},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
contents.append(Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
  color=COLORS["orange"],
  label="early, mid. ext.; N0",
))

scenario = Scenario(
  t_stages=["late"],
  midext=True,
  diagnosis={
    "ipsi": {"CT": {"III": True}},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
contents.append(Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
  color=COLORS["red"],
  label="adv., mid. ext.; ipsi LNL III",
))

draw(ax[0], contents, xlims=(0., 15.))
ax[0].legend(labelspacing=0.1)
ax[0].set_xlabel("risk [%]")
ax[0].set_yticks([])
ax[0].set_xticks(np.linspace(0, 15, 6))
ax[0].set_title("Contralateral LNL II")

# contra LNL III
contents = []

ax[1].set_yticks([])
ax[1].set_xticks(np.linspace(0, 15, 6))

ax[2].set_yticks([])
ax[2].set_xticks(np.linspace(0, 15, 6))

plt.savefig(paths.figure_dir / "fig-model-risks.png", dpi=400)
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-marginal-pdf.png", dpi=400)
plt.show()
```

>>>>

plt.savefig(paths.figure_dir / "fig-model-midext-evo.svg")
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-burnin-history.svg")
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-state-dist.svg")
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-prevalences-overall.svg")
plt.show()
```

@@

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[0,1],
  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[1,0],
  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
  locs=[(0.16, 0.99), (0.53, 0.55)],
)
split_legends(
  axes=axes[1,1],
  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
  locs=[(0.13, 0.55), (0.5, 0.99)],
)

plt.savefig(paths.figure_dir / "fig-model-prevalences-with-ipsi.svg")
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-prevalences-midext.svg")
plt.show()
```

@@

```{python}
import numpy as np
import h5py
import matplotlib.pyplot as plt
from matplotlib.legend_handler import HandlerTuple
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, draw, COLORS, split_legends
from scripts import paths, shared

@@

fig, axes = plt.subplots(nrows=nrows, ncols=ncols)

contents = {"II": [], "III": [], "IV": []}
mean_lists = {"II": [], "III": [], "IV": []}
defaults = {
  "density": True,
  "histtype": "stepfilled",
  "alpha": 0.7,
  "bins": 50,
}
t_stage_map = {"early": "early", "late": "adv."}
midext_map = {False: "lateralized", True: "mid. ext."}
colors = [COLORS["green"], COLORS["blue"], COLORS["orange"], COLORS["red"]]

def get_diag_label(diagnosis) -> str:
  label = ""
  is_n0 = True

  ipsi_positive = [lnl for lnl, status in diagnosis.get("ipsi", {}).get("CT", {}).items() if status]
  if len(ipsi_positive) > 0:
    is_n0 = False
    label += "i:" + ",".join(ipsi_positive)

  contra_positive = [lnl for lnl, status in diagnosis.get("contra", {}).get("CT", {}).items() if status]
  if len(contra_positive) > 0:
    is_n0 = False
    label += "; c:" + ",".join(contra_positive)

  return label if not is_n0 else "N0"

def get_label(scenario: Scenario) -> str:
  t = scenario.t_stages[0]
  m = scenario.midext
  d = eval(scenario.diagnosis)
  return f"{t_stage_map[t]}; {midext_map[m]}; {get_diag_label(d)}"

with h5py.File(name=paths.model_dir / "full" / "risks.hdf5", mode="r") as h5file:
  for dset in h5file.values():
    scenario = Scenario.from_dict(dict(dset.attrs))
    for_subplot = list(eval(scenario.involvement)["contra"].keys()).pop()
    mean_lists[for_subplot].append(dset[:].mean())

indices = {}
for lnl, means in mean_lists.items():
  indices[lnl] = np.argsort(means)

counter = {"II": 0, "III": 0, "IV": 0}
with h5py.File(name=paths.model_dir / "full" / "risks.hdf5", mode="r") as h5file:
  for dset in h5file.values():
    scenario = Scenario.from_dict(dict(dset.attrs))
    for_subplot = list(eval(scenario.involvement)["contra"].keys()).pop()
    c = counter[for_subplot]
    counter[for_subplot] += 1
    contents[for_subplot].append(
      Histogram(
        values=dset[:],
        kwargs={
          "label": get_label(scenario),
          "color": colors[indices[for_subplot][c]],
        },
      )
    )

xlims = {
  "II": (0, 15),
  "III": (0, 7),
  "IV": (0, 5),
}
for ax, (lnl, content) in zip(axes, contents.items()):
  draw(ax, content, xlims=xlims[lnl])
  ax.set_title(f"Contralateral LNL {lnl}")
  ax.set_yticks([])
  ax.set_xlabel("risk [%]")
  ax.legend(labelspacing=0.1)

axes[0].set_xticks(np.linspace(xlims["II"][0], xlims["II"][1], 6))

plt.savefig(paths.figure_dir / "fig-model-risks.svg")
plt.show()
```

@@

plt.savefig(paths.figure_dir / "fig-model-marginal-pdf.svg")
plt.show()
```
